---
title: "Shooting Occurrences in Toronto"
author: 
  - Varun Vijay
thanks: "Code and data are available at: https://github.com/Varun1005473462/starter_folder-main."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Statistics about shooting occurrences are a very important statistic to consider in Toronto as gun violence is a very serious problem in the modern world. I obtained and analyed yearly data from 2014 to 2019 about shooting occurences that were reported in Toronto from the city on a geographic basis. I found a mostly increasing trend in shooting occurences for Toronto from 2014 to 2019 and a close correlation between year and number of shooting occurences in the year. My findings have serious implications for the crime rate and the effectiveness of gun control in Toronto "
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

First paragraph is going to be motivational and broad based on the first sentence of abstract.

Second paragraph is about what was done and what was found. 

Third paragraph about implications

The remainder of this paper is: Section \@ref(data) explains data. Section \@ref(results) explains results, etc..... 



# Data

Paragraph or two introducing the dataset broadly. 

then show an extract of dataset (Table \@ref(tab:dataextract))

```{r dataextract, echo=FALSE}
library(tidyverse)
library(knitr)
shooting_occurences<-
  read_csv(here::here("inputs/data/shooting_occurences.csv"))
shooting_occurences %>% 
  select(OccurredYear,GeoDivision,Category, Count_) %>%
  slice(1:10) %>%
  kable(
    caption = "Table 1: First ten rows of a dataset that shows shooting Occurences in Toronto",
    col.names=c('Occured Year','Geo Division','Category','Count'),
    digits=1,
    booktabs=TRUE,
    linesep="")
```

Paragraph or two more about the Table \@ref(tab:dataextract)

Our dataset is of shooting occurrences in Toronto

I was interested in the relationship between the current year and the number of shooting occurrences that took place in said year because it was a measure of how effective gun control policies are in Toronto from year to year. I was also interested in how these shooting occurrences were split among the geographical divisions for each year included in the dataset.

Figure 1 shows the relationship between year and shooting occurrences and figure 2 shows how the geographical division of the shooting occurrences for each year.

```{r, echo=FALSE}
shooting_barplot <- 
  shooting_occurences %>% 
  mutate(Year = 
           case_when(OccurredYear==2014 ~ "2014",
                     OccurredYear==2015 ~ "2015",
                     OccurredYear==2016 ~ "2016",
                     OccurredYear==2017 ~ "2017",
                     OccurredYear==2018 ~ "2018",
                     OccurredYear==2019 ~ "2019",
                     ),
         Year = factor(Year,
                            levels = c("2014",
                                       "2015",
                                       "2016",
                                       "2017",
                                       "2018",
                                       "2019"
                                       )
                            )
         )
shooting_barplot %>%  
  ggplot(mapping = aes(x=OccurredYear,y=Count_)) +
  geom_bar(stat='identity')+
  labs(title='Figure 1')

shooting_barplot %>%  
  ggplot(mapping = aes(OccurredYear,Count_, fill=GeoDivision)) +
  geom_bar(stat='identity',position='dodge')+
  labs(title='Figure 2')

```



Talk more about it.



Condensing the data to find the maximum shooting occurences for each year
```{r, echo=FALSE}
shooting_occurences_2014<-shooting_occurences %>%
  filter(OccurredYear==2014)
count_2014<-sum(shooting_occurences_2014$Count_)
shooting_occurences_2015<-shooting_occurences %>%
  filter(OccurredYear==2015)
count_2015<-sum(shooting_occurences_2015$Count_)
shooting_occurences_2016<-shooting_occurences %>%
  filter(OccurredYear==2016)
count_2016<-sum(shooting_occurences_2016$Count_)
shooting_occurences_2017<-shooting_occurences %>%
  filter(OccurredYear==2017)
count_2017<-sum(shooting_occurences_2017$Count_)
shooting_occurences_2018<-shooting_occurences %>%
  filter(OccurredYear==2018)
count_2018<-sum(shooting_occurences_2018$Count_)
shooting_occurences_2019<-shooting_occurences %>%
  filter(OccurredYear==2019)
count_2019<-sum(shooting_occurences_2019$Count_)
Year<-c(2014,2015,2016,2017,2018,2019)
Shooting_count<-c(count_2014,count_2015,count_2016,count_2017,count_2018,count_2019)
shooting_occurences_year<-data.frame(Year,Shooting_count)
shooting_occurences_year
```


```{r, echo =FALSE}
library(modelsummary)
shooting_occurences_year %>%
  datasummary_skim()
```

```{r, echo =FALSE}
shooting_occurences_year %>%
  ggplot(mapping = aes(x = Year, y = Shooting_count)) +
  geom_point()+
  labs(title="Relationship between year and shooting occurences in that year")
```


```{r, echo=FALSE}
shooting_occurences_year %>%
  datasummary_correlation()
```


Talk way more about it. 


# Results

# Discussion

## First discussion point

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional details


\newpage


# References


